How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-04-09 docs/manuscript.Rmd
adding processing scripts for packages etc

```{r setupmain, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)

opts_chunk$set(echo = FALSE, warning = FALSE)
out_main_fig = TRUE

@@

@@

## Estimating travel time to closest ARMC {#methods-spatial}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).

Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale. We then compared three different metrics of access:

1) Travel time
2) Weighted travel time
3) Distance

extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell. We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

@@

For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for this, we estimated clinic level undersubmission as the proportion of days on which forms were submitted in a given year, excluding any periods for which there were no forms submitted for 15 consecutive days. We then divided the total number of bites reported to each district at each clinic in a given year by the undersubmission rate for that clinic in that year. Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates. The data from the Moramanga district were completely reported across the 31 month period.

@@

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) as follows: 
$$ \mu = e^{(\beta_{t}T \ + \ \beta_0 \ )}P$$
where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T$) and the human population size ($P$) (an offset which scales the predicted number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu = \sum \limits_{j=1}^je^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$

As travel times are correlated with human population size (Figure S3.1A), we also compared how well bites were predicted by human population size alone, and in combination with travel times. For the models with population size, we used a model framework with either population size alone ($\mu = e^{(\beta_pP + \beta_0 )}$) or population size and travel times ($\mu = e^{(\beta_{t}T + \beta_pP + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.

For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercpet).

To test our models predictions further, we used parameter estimates from each model to generate predictions for data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data).

Since we're assuming for poisson errors: we tested for overdispersion.

Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to uncorrected data.

@@

<!-- TO DO: simplify discussion here but bring in some figs, etc to Supplementary section to explain assumptions! -->

We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2). We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$). We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons. As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and generates similar human exposure incidences as observed previously across Africa [cite here!] . $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.

```{r fig 2, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M2.jpeg"))
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. All parameter values and sources are listed in Table 1.*

$B_i$, the number of total reported bites, is drawn from a poisson distribution with the mean predicted number of bites predicted from the models of incidence as a function of travel times. To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. We draw $p_{rabid}$ from a uniform distribution ranging from 0.2 - 0.6 based on investigations of bite patients reporting in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2018]. So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
where $\rho_{max}$ is the maximum reporting. We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018]. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies. 
We subtract $R_i$ from $E_i$ to get the number of unreported exposures ($U_i$). Deaths due to rabies in each administrative unit ($D_i$) are then estimated as:
$$ D_i = binom(U_i,\ p_{death}) $$
Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_i = binom(R_i, \ p_{death})$). Table 1 lists all parameter values and sources used in the decision tree. \

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@whorabiesmodelingconsortiumPotentialImpactImproved2019] |
| $E_{min} − E_{max}$ |	11.1 - 55.7 per 100,000 persons	| Range of human exposure incidence calculated from an assumed dog rabies incidence of 1%, human:dog ratios between 7 - 35, and $p_{exp}$. | [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
| $p_{rabid}$	| 0.2 - 0.6 |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

@@

```{r ncsbs}
## candidate points
csbs <- read.csv(here("data/raw/csbs.csv", stringsAsFactors = FALSE))
csbs %>% 
  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
```

@@

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric weighted by the reduction in travel times, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). We also explore the scenario in which all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning. Resulting travel time estimates and locations of when and where clinics were added are shown in [Supplementary section S5](#S5). Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites and then aggregate these to the clinic catchment. As clinics are added, multiple clinics may serve a single district, but in general, the majority of the population within a given commune are closest to a single clinic (Fig S5.3B). Therefore, we assign the catchment clinic as the one which is the closest for the majority of people living in a commune. For the district model, we disaggregate the reported bites to the commune level (assuming uniform incidence within the district). For each catchment, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e. wastage due to injection error). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
We estimate burden across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden. Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, $p_{rabid}$ at 0.2 and 0.6, and $\rho_{max}$ at 0.85 and 0.99 to obtain maximum and minimum estimates of burden. We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplementary section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.

## Data availability
All analyses were done in R and using these packages (list packages with commas and citations). All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo. The raw bite patient data at the national level is maintained by IPM. This database was last queried on *date here* for these analyses. Anonymized raw data from this database are available upon request from IPM. Anonymized raw data collected from the Moramanga District are shared in the archived repository.

@@

## Estimating travel times to the closest ARMC
```{r ttime stats}
ttime_stats <- read.csv(here("output/stats/prop_pop_ttimes.csv"))
# get prop of pop > 3 hrs away at baseline
ttime_stats %>%
  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
  sum(prop_pop) -> prop_over3_base
ttime_stats %>%
  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
  sum(prop_pop) -> prop_over6_base
```

@@

```{r fig M2, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M3_ttimes.jpeg"))
```
*__Figure 3.A)__ Travel times to the closest ARMC estimated at an ~ 1km^2 scale for Madagascar. __B)__ Distribution of the population across travel times. __C)__Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level. For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune. The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times ~ estimated travel times for each data source are also shown. The dashed black line is the 1:1 line.*

## Estimating average annual bite incidence at the district and commune level
We assigned each district or commune to a catchment based on which clinic was the closest for the majority of the population. While there are discrepancies between commune and district catchment assignments (Fig S2.1A), > 75% of the population in a given district or commune were closest to a single clinic (Fig S2.B). In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A) and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).

@@

Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1). However, we found that corrected estimates of reproted bites generated predictions of vial demand close to what was provisioned over the same time period ([Figure S2.2A and B](#figS2.2)). We excluded any districts in a catchment of a clinic which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, corresponding to `r length(mada_districts) - nrow(district_bites)` districts, in grey in Fig 3A). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).

@@

```{r}
outliers <- read.csv(here("output/stats/outliers.csv"))
head(outliers)
```

Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship. There were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher incidence at low travel times at the commune level compared to the district.

## Modeling reported bite incidence as a function of travel times {#results-bitemods}

All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2). Population size alone was the poorest fit to the data as estimated by DIC, and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3). Sensitivity to underreporting. For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2). In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Fig 3). These predictions were robust to how we corrected for incomplete submission of data by clinics ([supplementary section S4](#S4)). As the random intercept models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence. (OD COMP--use the Moramanga model with OD & the District model with OD)

```{r fig 3, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M3.jpeg"))
```
_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons. The estimated relationship between travel time in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis). The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept. The envelopes shows the 95% CI given the random intercept for the commune and district models. The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept. The points show the data used to fit the models (National data (triangles) at the district level used to fit the District and Commune models, and Moramanga data (circles) at the commune level used to fit the Moramanga model)._

>>>>

```{r setupmain, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)
library(magrittr)

opts_chunk$set(echo = FALSE, warning = FALSE)

@@

trans_perc <- function(x) {
  round(x, 2)*100
}

@@

## Estimating travel time to closest ARMC {#methods-spatial}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).  
Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale. We then extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell. We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

@@

For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1). We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period. We did not have the date or month when vials were provisioned, so we compared the totals across this three year period.

We then divided the total number of bites reported to each clinic from each clinic in a given year by the undersubmission rate for eachclinic. Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level. The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period.

@@

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$
where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the predicted number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$

As travel times are correlated with human population size (Figure S3.1), we also compared how well bites were predicted by human population size alone, and in combination with travel times. For the models with population size, we used a model framework with either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.

For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercept). We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by comparing the ratio of the sum of squared residuals to the expected $\chi^2_{n - k}$ distribution given a poisson error structure [@gelman2006data].

The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment. We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book).

To test our models predictions further, we used parameter estimates from each model to generate predictions for data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data). Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to the raw data uncorrected for submission (i.e. assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs. 15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).

@@

We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2). As with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], to model uncertainty in parameter estimates we used triangular distributions for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures. The range for these distributions was taken from a previous study in Moramanga and also informed by recent methods for estimating underlying rabies incidence in an endemic setting when data is lacking([@whorabiesmodelingconsortiumPotentialImpactImproved2019]).

Specifically: annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination . To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$). We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons. As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and generates similar human exposure incidences as observed previously across Africa [cite here!] . $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.

```{r fig 1, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M1_sankey.tiff"))
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. Both $E_i$ and $p_{rabid}$ are drawn from a triangular distribution; all parameter values and sources are listed in Table 1.*

$B_i$, the number of total reported bites, is the mean predicted number of bites from the models of bite incidence as a function of travel times. To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures, drawn from a triangular distribution with range between 0.2 - 0.6. So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
where $\rho_{max}$ is the maximum reporting. We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018]. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies. 
We subtract $R_i$ from $E_i$, the number of total human rabies exposures drawn from a triangular distribution with range of 15.6 - 76 exposures per 100,000 persons, to get the number of unreported exposures ($U_i$). Deaths due to rabies in each administrative unit ($D_i$) are then estimated by mutliplying $D_i$ times $p_{death}$.
Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_ip_{death}$). Table 1 lists all parameter values and sources used in the decision tree. \

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $E_i$ |	Triangular(a = 15.6 b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see S4|
| $p_{rabid}$	| Triangular(a = 0.2 b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

@@

```{r ncsbs}
## candidate points
csbs <- read.csv(here("data/raw/csbs.csv"), stringsAsFactors = FALSE)
csbs %>% 
  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
```

@@

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). We also explore the scenario in which all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning. Resulting travel time estimates and locations of when and where clinics were added are shown in [Supplementary section S5](#S5). Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level. As clinics are added, multiple clinics may serve a single district or commune (Fig TO CHECK S5.3B). Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest (Fig CHECK maps). For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e. wastage due to injection error). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and models of burden. Additionally, we also examine the systematic variation in human rabies exposure incidence with human population (a potential proxy for changes in the dog population, see section S6) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size. 
We also generate scaling factors to look at how systematic variation in rabies incidence with human population size could affect our estimates of burden (Fig S6.1). Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e. if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases. Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e. if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.

## Data and code 
All analyses were done in R and using these packages (list packages with commas and citations). All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo. The raw bite patient data at the national level is maintained by IPM. This database was last queried on *date here* for these analyses. Anonymized raw data from this database are available upon request from IPM. Anonymized raw data collected from the Moramanga District are shared in the archived repository.

@@

## Estimating travel times to the closest ARMC {#results-ttimes}
```{r ttime stats}
ttime_stats <- read.csv(here("output/stats/prop_pop_ttimes.csv"))
# get prop of pop > 3 hrs away at baseline
ttime_stats %>%
  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3"))) %$%
  sum(prop_pop) -> prop_over3_base
ttime_stats %>%
  filter(scenario == "baseline", !(cut_times %in% c("0-1", "1-2", "2-3", "4-5", "5-6"))) %$%
  sum(prop_pop) -> prop_over6_base
```

@@

```{r fig M2, fig.fullwidth=TRUE}
include_graphics(here("figs/main/M2_ttimes.jpeg"))
```
*__Figure 3.A)__ Travel times to the closest ARMC estimated at an ~ 1km^2 scale for Madagascar. __B)__ Distribution of the population across travel times. __C)__Correlation between patient reported travel times to the Moramanga ARMC and travel times weighted by population (mean at the commune level) and between reported driving times between GPS points and travel time estimates at the grid cell level. For the commune means, the vertical point lineranges show the 95% quantiles for reported travel times, and the point size shows the number of observations for each commune. The best fit lines (pink and purple lines) and 95% confidence interval (shading) from a linear model where observed travel times ~ estimated travel times for each data source are also shown. The dashed black line is the 1:1 line.*

## Estimating average annual bite incidence at the district and commune level {#results-biteinc}
We assigned each district or commune to a catchment based on which clinic was the closest for the majority of the population. While there are discrepancies between commune and district catchment assignments (Fig S2.1A), > 75% of the population in a given district or commune were closest to a single clinic (Fig S2.B). In addition, for the national data we found that for most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A), and the majority of bites reported to a clinic came from within the catchment assigned by the travel time metric (Fig S2.1C).

@@

Clinic submission of forms was highly variable over the period covered by the dataset (Figure S2.1). We found that estimates of vial demand based on uncorrected bite patient numbers are generally lower than the true number of vials provisioned for those clinics with substantial undersubmission of forms ([Figure S2.1B](#figS2.1)), and correcting patient numbers by under-submission of forms based on the 15 day consecutive day cut-off results in estimates of vial demand closer to the observed vials provisioned ([Figure S2.2A and B](#figS2.2)). We excluded any clinics which submitted less than 10 forms (excluded `r nrow(ctar_metadata) - max(district_bites$catch)` catchments, in grey in Fig 3A). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total (Fig 4A), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 4B).

@@

```{r}
outliers <- read.csv(here("output/stats/outliers.csv"))
```

Bite incidence estimates generally decreased with weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship. For the national data, there were also two outliers, Toamasina II (the sub-urban commune surrounding the city of Toamasina) and Soanierana Ivongo, which both had higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data when aggregated to the district fell within the ranges observed from the National dataset, at the commune scale they showed more variation, with higher reported incidence at low travel times at the commune level compared to the district.

## Modeling reported bite incidence as a function of travel times {#results-bitemods}
All models and posterior estimates of parameters converged ({inline stat for max psrf & mpsrf}). Population size was the poorest predictor of bites (Fig S3.2, Table S3.1) and adding population size to the model did not improve the fit to the national level data (Figure S3.2), although these models did have lower DIC values for the Moramanga data (Table S3.1). However, when used to predict data that were not used to fit the models (i.e. model estimates from fitting to the Moramanga data were used to predict the national data and vice versa), only models with incidence offset by population size (flatPop) rather than scaling with population size (addPop) generated reasonable predictions to the data (Figure S3.3).\

All of the candidate models which included travel times produced similar predictions when compared to the observed data (Fig S3.2). Population size alone was the poorest fit to the data as estimated by DIC (Table S3.1), and models with population size as an additional covariate did not generate realistic predictions to the observed data or when used to predict out of fit (Figs S3.2 and S3.3). For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Fig S3.2). These predictions were robust to how we corrected for incomplete submission of data by clinics (Fig S3.5). The models without the random effect were significantly overdispersed, but the catchment random effect still captured additional variation beyond this overdispersion (Fig S3.4). The parameter estimates from the different models were similar, with the model fit to the Moramanga data showing the highest impact of travel times and the district model the lowest (Fig 5B-D), however the increased bite incidence observed at low travel times in the Moramanga data was not captured by the commune model (Fig 5A). To capture the range of estimates of travel time impact, we compared the district model fitted to the national data with the commune scale model fitted to the Moramanga data in subsequent analyses.
